JEThread: A Thread Pool Implementation in C

### 2018-05-04 thread-pool-implementation/paper.tex

\subsection{Work Stealing}
One of our goals is to achieve automatic load balancing. For this purpose, we implement work stealing for each worker thread. For each worker, we first do a sem\_trywait() to see if there is any job on its private queue. If there is none, then it randomly selects a victim thread and tries to steal a task from the victim and executes it. If keeps doing so until there is no more work to steal or if it has stolen $N$ times, where $N$ is the number of workers. This is our first attempt at work stealing. To optimize this, we could have kept randomly selecting victims until there is a victim with work above a certain threshold or up to a certain number of tries. To avoid all workers trying to steal when there is no tasks available, we can also use randomization to determine if a worker should try work stealing or just block until it is assigned a job. Another optimization we can perform is to steal more than one work at a time since we want it to be worth the overhead of running the algorithm. We evaluated the impact of the current work stealing techniques and found that even the naive approach improves JEThread's performance slightly compared to the implementation without work stealing.

\subsection{Internal Synchronization}
Each thread owns a private task queue to store all its assigned tasks. Each queue is protected by either a mutex or a spinlock depending on users' specification. However, it is possible that a thread's task queue is empty when it is first initialized. It will waste CPU resource if such task is kept up and running without executing any tasks. Thus, we use semaphore to improve the efficiency of the thread pool. We let the main thread invoking sema\_post on each thread once it finishes adding tasks to the thread's task queue to indicate the availability of tasks. However, the thread's task queue can be empty for other reasons. The most important one is that the thread finishes executing all assigned tasks. Thus, we implement work stealing algorithm to keep the idle thread busy by letting it try to steal tasks from other threads by randomly choosing one victim. Since we don't want tasks stealing to create too much overhead as it tries to grab tasks from other threads, we simply let the idle thread using pthread\_mutex\_trylock() or pthread\_spin\_trylock(). If the stealing succeeds, the tasks is grabbed from the victim's task queue and executed. Otherwise, the steal will fail and the failure will be reported to the calling thread. If the steal fails too many times, that indicates many threads' task queue is either already empty and the thread pool is getting close to running out of tasks, or every victim is holding the lock. However, the possibility for the every victim to hold the thread just as the calling thread is trying to grab it is diminishing when more stealing are attempted. In this case, letting more threads trying to steal tasks will create unnecessary overhead, thus the thread will be blocked to prevent wasting resources. If eventually the threads run out of tasks, it will increase \textbf{wait\_sema} to let the main thread know that it finishes all its job. If all threads signify that they've finihsed their tasks, then thread pool is terminated.

@@

There are several goals we are aiming to achieve in the experiment. First, we want to run both long and short tasks in the threal pool to analyze which type of tests have the most impact on the performance on the program. We've created various tests for the thread pool to perform and we will show the data collected later in the analysis section. We hope the low overhead of switch between tasks will help us outperform thread-based implementation as well as go-based implementation. In addition, we implement work stealing mechanism to improve the work throughput done by the thread pool. We hope the work stealing function will further improve the efficiency of the thread pool.

We create same tasks for every implementation to execute to ensure fairness. We've selected three types of tests that represent three typical scenarios in most applications. We designed a task that accesses file system and performs read and write operations. Since accessing file system usually requires blocking which generates large overhead, this task is perfect to represent tasks that takes more than a few CPU cycles to finish. We also designed a task that take in an argument, and the computer then performs sumation based on the input and generates a simple hash value. This represents a short to medium tasks present in most applications and we selected it as a candidate to represent short tasks. We also include a task that does nothing in order to test sheer performance of the thread pool without the influence of any outside factors. In addition, we mixed the number of long and short tasks in each benchmark in order to generalize the real life scenario where a mixture of different tasks are executed. We want to show that the speedup is not or at least minimally affected by the type of tasks executed. Then, we test the scalability of the thread pool itself by gradually increase the number of threads in the thread pool running $10,000$ tasks. The result matches with our expectation as the scalability is achieved when increasing the number of threads in the thread pool.\\

@@

\section{Conclusion}
	Theoratically, thread pool can be very efficient especially dealing with a large number of relatively short tasks concurrently. In order to acheive maximum performance boost, the use of context switch should be minimized. The thread pool does a greate job of reducing the overhead in context switch. In addition, work stealing further facilitates the work throughput by letting idle threads execute other threads' tasks. However, thread pool is not a silver bullet that solves every problem of thread. The overhead of thread context switch is transfered to each thread's task queue as executing tasks requires dequeing the task each time, which also applies to job stealing. We've also conducted experiement with creating only one task queue shared between all threads, but the same thing still applies here that the overhead is transfered to the queue since accessing tasks is performed with lock's protection. In addition, round-robin style task distribution may not be the best way to distribute tasks as calling thread is unware of the actual workload in its distribution target. The work stealing is also very tricky to implement as the stealer is not aware of the workload distribution during runtime. If the majority of threads finish execution while a few survivor threads with many tasks are still alive, the parallelism can be compromised since a few threads become the bottleneck that causes main threads waiting unnecessarily. Additional improvements such that runtime dynamic workload check and compiler support for user-guided optimization will further increase the usibility of thread pool.\\

	In summary, the thread pool is great for running a large amount of tasks effciently with relatively small number of threads. In parallel applications with many short tasks to perform thread will be a strong candidate solution.

>>>>

\subsection{Work Stealing}
One of our goals is to achieve automatic load balancing. For this purpose, we implement work stealing for each worker thread. For each worker, we first do a sem\_trywait() to see if there is any job on its private queue. If there is none, then it randomly selects a victim thread. Since we don't want work stealing to create too much overhead as it tries to grab tasks from other threads, we first let the stealer try peeking at the victim's queue using pthread\_mutex\_trylock() or pthread\_spin\_trylock(). If the stealing succeeds, a task is grabbed from the victim's task queue and executed. Otherwise, the steal will fail. If the steal fails too many times, that indicates many threads' task queue is already empty and the thread pool is getting close to running out of tasks. In this case, letting more threads trying to steal tasks will create unnecessary overhead, thus the thread will be blocked to prevent wasting resources.

This is our first attempt at work stealing. To optimize this, we could have kept randomly selecting victims until there is a victim with work above a certain threshold or up to a certain number of tries. To avoid all workers trying to steal when there is no tasks available, we can also use randomization to determine if a worker should try work stealing or just block until it is assigned a job. Another optimization we can perform is to steal more than one work at a time since we want it to be worth the overhead of running the algorithm. Although not reflected in the evaluation, we did try to evaluate the impact of the current work stealing techniques and found that even the naive approach improves JEThread's performance slightly.

\subsection{Internal Synchronization}
Each worker thread owns a private task queue to store all its assigned tasks. Each queue is protected by either a mutex or a spinlock depending on users' specification. We highly recommend using a spinlock because we've ran experiments for both types of locks and observed that the spinlock always provides better performance. This is because the critical section is small and the code design rarely leads to lock contention.

In addition to preserving correctness through protected access to each worker's queue, we also try to keep JEThread's overhead low through efficient implementation. For example, when a worker runs out of tasks to execute, it can either busy wait or block. Because busy waiting will exhaust CPU resource, we decide to use a semaphore \textbf{task\_sema} to block a worker while there is no available task. When the main thread assigns tasks to the worker thread, it calls sema\_post on the worker's task\_sema.

In order to wait for all worker threads to finish, in thread\_pool\_wait() we first mark each worker's private state \textbf{count} to true, then wake up each worker using \textbf{task\_sema} , and wait on each worker's \textbf{wait\_sema}. Once a worker wakes up, if there is no more job on its private queue and \textbf{count} is true, it wakes up the main thread to indicate that it has finished the current set of tasks.

@@

There are several goals we are aiming to achieve in the experiment. Firstly, we test the flexibility of the thread pool by running a mixture of long(blocking) and short tasks and evaluate how we perform compared to other implementations. Secondly, we evaluate the overhead of JEThread by running a mixture of short and empty tasks. The tasks we ran are common tasks in most web servers. The long task accesses file system, creates a file and writes 1106 bytes to the file. Since accessing file system usually requires blocking which generates large overhead, this task is perfect to represent tasks that takes lots of CPU cycles to finish. The short task generates a simple hash value and the empty task does nothing in order to test sheer overhead of the thread pool without the influence of any outside factors.

To show flexibility, we show that JEThread is able to outperform at least two other implementations despite of the workloads. We compared JEThread with pthread per core, pthread per task and Goroutines. Then, we test the scalability of JEThread by gradually increase the number of threads in the thread pool running $10,000$ short tasks. The result matches with our expectation as the scalability is achieved when increasing the number of threads in the thread pool.\\

@@

\section{Conclusion}
	A thread pool should be efficient when managing a large number of tasks and executing them concurrently. In JEThread, we use work stealing to facilitate the work throughput by letting idle threads execute other threads' tasks. However, thread pool is not a silver bullet that solves every problem of thread. The overhead of thread context switch is transferred to each thread's task queue as executing tasks requires dequeueing the task each time. We've also conducted experiment with creating only one task queue shared between all threads, but the same thing still applies here that the overhead is transferred to the queue since accessing tasks is performed with lock's protection. In addition, round-robin style task distribution may not be the best way to distribute tasks as calling thread is unaware of the actual workload in its distribution target. The work stealing is also very tricky to implement as the stealer is not aware of the workload distribution during runtime. If the majority of threads finish execution while a few survivor threads with many tasks are still alive, the parallelism can be compromised since a few threads become the bottleneck and cause the main thread waiting unnecessarily. Additional optimization such as runtime dynamic workload check and compiler support for user-guided optimization will further increase the usability of thread pool.\\

	In summary, the thread pool is great for running a large amount of tasks efficiently with relatively small number of threads. In parallel applications with many short tasks to perform thread will be a strong candidate solution.
